---
title: "Arbacia lixula differential expression analysis"
author: "Vanessa Arranz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
here::i_am("analysis/01_deseq2_analysis.Rmd")
```

# Purpose

This analysis reproduces the DESeq2 comparisons reported in Arranz et al.
(2026). The model contains one three-level treatment factor:

- `CC`: control-origin individuals maintained under control conditions.
- `CV`: control-origin individuals exposed to vent conditions.
- `VV`: vent-origin individuals maintained under vent conditions.

The three contrasts represent the chronic response (`VV vs CC`), acute
plastic response (`CV vs CC`), and genotype/origin effect under shared vent
conditions (`VV vs CV`). Positive log2 fold changes indicate higher expression
in the first group named in each contrast.

# Packages and paths

```{r packages}
library(DESeq2)
library(dplyr)
library(ggplot2)
library(here)
library(pheatmap)
library(readr)
library(RColorBrewer)
library(tibble)
```

```{r paths}
count_file <- here("data", "count_curated.csv")
metadata_file <- here("data", "sample_metadata.csv")

results_dir <- here("results", "deseq2")
figure_dir <- here("results", "figures", "deseq2")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

stopifnot(file.exists(count_file), file.exists(metadata_file))
```

# Import and validate data

```{r import}
count_table <- read_csv(count_file, show_col_types = FALSE)
metadata <- read_csv(metadata_file, show_col_types = FALSE)

required_metadata <- c(
  "sample_id", "treatment", "origin_environment",
  "experimental_environment"
)
if (!all(required_metadata %in% names(metadata))) {
  stop("sample_metadata.csv is missing one or more required columns.")
}
if (!"Geneid" %in% names(count_table)) {
  stop("count_curated.csv must contain a 'Geneid' column.")
}
if (anyDuplicated(count_table$Geneid)) {
  stop("Duplicated gene identifiers detected in count_curated.csv.")
}
if (anyDuplicated(metadata$sample_id)) {
  stop("Duplicated sample identifiers detected in sample_metadata.csv.")
}

count_data <- count_table |>
  column_to_rownames("Geneid") |>
  as.matrix()

storage.mode(count_data) <- "numeric"
if (anyNA(count_data) || any(count_data < 0) || any(count_data %% 1 != 0)) {
  stop("The count matrix must contain non-negative integer counts without NAs.")
}

if (!setequal(colnames(count_data), metadata$sample_id)) {
  missing_metadata <- setdiff(colnames(count_data), metadata$sample_id)
  missing_counts <- setdiff(metadata$sample_id, colnames(count_data))
  stop(
    "Sample mismatch. Missing from metadata: ",
    paste(missing_metadata, collapse = ", "),
    "; missing from counts: ",
    paste(missing_counts, collapse = ", ")
  )
}

metadata <- metadata |>
  mutate(treatment = factor(treatment, levels = c("CC", "CV", "VV"))) |>
  slice(match(colnames(count_data), sample_id)) |>
  as.data.frame()
rownames(metadata) <- metadata$sample_id

stopifnot(identical(colnames(count_data), rownames(metadata)))

dim(count_data)
table(metadata$treatment)
```

# Construct the DESeq2 dataset

```{r dds}
dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData = metadata,
  design = ~ treatment
)

# Low-expression filter reported in the associated publication: retain genes
# with at least 10 counts in at least 4 samples.
keep <- rowSums(counts(dds) >= 10) >= 4
dds <- dds[keep, ]

cat("Genes in original matrix:", nrow(count_data), "\n")
cat("Low-expression genes removed:", sum(!keep), "\n")
cat("Genes retained for DESeq2:", nrow(dds), "\n")

stopifnot(nrow(dds) == 17449)
```

# Exploratory analysis

```{r transformations}
dds <- estimateSizeFactors(dds)
vst_data <- vst(dds, blind = FALSE)
```

```{r sample-distance, fig.width=8, fig.height=7}
sample_distances <- dist(t(assay(vst_data)))
sample_distance_matrix <- as.matrix(sample_distances)
rownames(sample_distance_matrix) <- paste(
  colData(vst_data)$treatment,
  colnames(vst_data),
  sep = " – "
)
colnames(sample_distance_matrix) <- rownames(sample_distance_matrix)

distance_colours <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
pheatmap(
  sample_distance_matrix,
  clustering_distance_rows = sample_distances,
  clustering_distance_cols = sample_distances,
  col = distance_colours,
  border_color = NA
)
```

```{r pca, fig.width=7, fig.height=5}
treatment_colours <- c(CC = "#65b3c1", CV = "#f09e3a", VV = "#e36348")

pca_plot <- plotPCA(vst_data, intgroup = "treatment") +
  scale_colour_manual(values = treatment_colours) +
  theme_classic(base_size = 13) +
  labs(colour = "Treatment")

pca_plot
ggsave(
  file.path(figure_dir, "pca_all_genes.pdf"),
  pca_plot,
  width = 7,
  height = 5
)
```

# Differential expression

```{r run-deseq2}
dds <- DESeq(dds)

contrast_definitions <- list(
  VV_vs_CC = c("treatment", "VV", "CC"),
  CV_vs_CC = c("treatment", "CV", "CC"),
  VV_vs_CV = c("treatment", "VV", "CV")
)

get_contrast <- function(dds_object, contrast_vector, alpha = 0.05) {
  results(dds_object, contrast = contrast_vector, alpha = alpha) |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    arrange(padj)
}

contrast_results <- lapply(
  contrast_definitions,
  function(x) get_contrast(dds, x)
)

invisible(
  Map(
    function(result_table, contrast_name) {
      write_csv(
        result_table,
        file.path(results_dir, paste0("deseq2_", contrast_name, ".csv")),
        na = ""
      )
    },
    contrast_results,
    names(contrast_results)
  )
)
```

```{r summary}
summarise_contrast <- function(result_table) {
  tibble(
    tested_genes = sum(!is.na(result_table$padj)),
    significant_padj_0.05 = sum(result_table$padj < 0.05, na.rm = TRUE),
    up_lfc1 = sum(
      result_table$padj < 0.05 & result_table$log2FoldChange > 1,
      na.rm = TRUE
    ),
    down_lfc1 = sum(
      result_table$padj < 0.05 & result_table$log2FoldChange < -1,
      na.rm = TRUE
    )
  )
}

deg_summary <- bind_rows(
  lapply(contrast_results, summarise_contrast),
  .id = "contrast"
)
write_csv(deg_summary, file.path(results_dir, "differential_expression_summary.csv"))
deg_summary
```

# Export reusable objects

```{r export-objects}
saveRDS(dds, file.path(results_dir, "dds.rds"))
saveRDS(vst_data, file.path(results_dir, "vst_data.rds"))

writeLines(
  rownames(dds),
  file.path(results_dir, "tested_gene_universe.txt")
)
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
