---
title: "Functional annotation and GO enrichment with topGO"
author: "Vanessa Arranz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
here::i_am("analysis/02_go_enrichment_topgo.Rmd")
```

# Purpose

Gene Ontology annotations were assigned to the longest transcript of each
predicted *Arbacia lixula* protein-coding gene using eggNOG-mapper v2. This
analysis uses the resulting gene-to-GO associations and applies the `elim`
algorithm and Fisher's exact test in topGO to
up- and down-regulated genes from each DESeq2 contrast. As in the original
analysis, enrichment gene sets use `padj < 0.05` and
`abs(log2FoldChange) > 1`.

# Packages and paths

```{r packages}
library(dplyr)
library(here)
library(readr)
library(tidyr)
library(topGO)
```

```{r paths}
deseq_dir <- here("results", "deseq2")
association_file <- here("data", "annotations", "gene2go.tsv")
results_dir <- here("results", "topgo")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(association_file)) {
  stop(
    "Missing data/annotations/gene2go.tsv. The file must contain two ",
    "tab-separated columns: gene_id and semicolon-separated GO terms."
  )
}
```

# Read gene-to-GO associations

```{r gene2go}
associations <- read_tsv(
  association_file,
  col_names = FALSE,
  comment = "#",
  show_col_types = FALSE
) |>
  select(1, 2) |>
  setNames(c("gene_id", "GO")) |>
  filter(
    !tolower(gene_id) %in% c("gene", "gene_id", "geneid", "query")
  ) |>
  filter(
    !is.na(gene_id), gene_id != "",
    !is.na(GO), GO != "", GO != "-"
  ) |>
  separate_rows(GO, sep = "[;,]") |>
  mutate(GO = trimws(GO)) |>
  filter(grepl("^GO:[0-9]+$", GO)) |>
  distinct(gene_id, GO)

gene2GO <- split(associations$GO, associations$gene_id)
cat("Genes with GO annotations:", length(gene2GO), "\n")
```

# Functions

```{r functions}
run_topgo_ontology <- function(gene_universe, selected_genes, ontology) {
  gene_vector <- factor(as.integer(gene_universe %in% selected_genes))
  names(gene_vector) <- gene_universe

  go_data <- new(
    "topGOdata",
    description = paste("Arbacia lixula", ontology),
    ontology = ontology,
    allGenes = gene_vector,
    geneSelectionFun = function(x) x == 1,
    annot = annFUN.gene2GO,
    gene2GO = gene2GO
  )

  fisher_result <- runTest(
    go_data,
    algorithm = "elim",
    statistic = "fisher"
  )

  result_table <- GenTable(
    go_data,
    elimFisher = fisher_result,
    orderBy = "elimFisher",
    topNodes = length(usedGO(go_data))
  )

  result_table |>
    mutate(
      ontology = recode(
        ontology,
        BP = "Biological Process",
        MF = "Molecular Function",
        CC = "Cellular Component"
      ),
      elimFisher_numeric = suppressWarnings(
        as.numeric(sub("^< ", "", elimFisher))
      )
    )
}

run_topgo_set <- function(result_table, direction) {
  tested <- result_table |>
    filter(!is.na(padj)) |>
    pull(gene_id)

  selected <- result_table |>
    filter(
      !is.na(padj),
      padj < 0.05,
      if (direction == "up") log2FoldChange > 1 else log2FoldChange < -1
    ) |>
    pull(gene_id)

  # topGO can only use genes with at least one GO association.
  universe <- intersect(tested, names(gene2GO))
  selected <- intersect(selected, universe)

  if (length(selected) == 0) {
    warning("No selected annotated genes for direction: ", direction)
    return(tibble())
  }

  bind_rows(lapply(c("BP", "MF", "CC"), function(ontology) {
    run_topgo_ontology(universe, selected, ontology)
  })) |>
    mutate(
      direction = direction,
      universe_size = length(universe),
      selected_genes = length(selected)
    )
}
```

# Run all contrasts

```{r enrichment}
contrast_files <- c(
  VV_vs_CC = "deseq2_VV_vs_CC.csv",
  CV_vs_CC = "deseq2_CV_vs_CC.csv",
  VV_vs_CV = "deseq2_VV_vs_CV.csv"
)

all_enrichment <- lapply(names(contrast_files), function(contrast_name) {
  input_file <- file.path(deseq_dir, contrast_files[[contrast_name]])
  if (!file.exists(input_file)) {
    stop("Run analysis/01_deseq2_analysis.Rmd first: ", input_file)
  }

  result_table <- read_csv(input_file, show_col_types = FALSE)

  bind_rows(
    run_topgo_set(result_table, "up"),
    run_topgo_set(result_table, "down")
  ) |>
    mutate(contrast = contrast_name, .before = 1)
}) |>
  bind_rows()

write_csv(
  all_enrichment,
  file.path(results_dir, "topgo_all_contrasts.csv"),
  na = ""
)

significant_enrichment <- all_enrichment |>
  filter(elimFisher_numeric < 0.05)

write_csv(
  significant_enrichment,
  file.path(results_dir, "topgo_significant_p0.05.csv"),
  na = ""
)

significant_enrichment
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
