---
title: "Identification of biomineralization-related genes"
author: "Vanessa Arranz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
here::i_am("analysis/03_biomineralization_genes.Rmd")
```

# Purpose

This analysis links *Arbacia lixula* eggNOG-mapper annotations to two published
sea-urchin skeletogenic/biomineralization gene lists. Identifiers are
standardised before joining, duplicate reference identifiers are removed, and
one best eggNOG hit is retained per *A. lixula* gene.

# Packages and paths

```{r packages}
library(dplyr)
library(here)
library(readr)
library(readxl)
library(stringr)
```

```{r paths}
annotation_dir <- here("data", "annotations")

# Preferred input: the unmerged eggNOG-mapper annotation table.
eggnog_file <- here(annotation_dir, "eggnog_annotations.csv")

# Compatibility input: the previously generated merged table can be used when
# the original eggNOG table is unavailable.
legacy_merged_file <- here(annotation_dir, "eggnog_biomineralization.csv")

dylus_file <- here(annotation_dir, "skeletogenic_genes_dylus.tsv")
pespeni_file <- here(annotation_dir, "biomineralization_genes_pespeni.xlsx")

results_dir <- here("results", "annotations")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

stopifnot(file.exists(dylus_file), file.exists(pespeni_file))
if (!file.exists(eggnog_file) && !file.exists(legacy_merged_file)) {
  stop(
    "Provide either data/annotations/eggnog_annotations.csv or ",
    "data/annotations/eggnog_biomineralization.csv."
  )
}
```

# Import reference gene lists

```{r reference-lists}
normalise_spu_id <- function(x) toupper(trimws(as.character(x)))

dylus <- read_tsv(dylus_file, show_col_types = FALSE) |>
  mutate(spu_id = normalise_spu_id(spu_id)) |>
  filter(!is.na(spu_id), spu_id != "") |>
  distinct(spu_id, .keep_all = TRUE)

pespeni <- read_excel(pespeni_file) |>
  mutate(spu_id = normalise_spu_id(spu_id)) |>
  filter(!is.na(spu_id), spu_id != "") |>
  distinct(spu_id, .keep_all = TRUE)

cat("Unique Dylus SPU identifiers:", nrow(dylus), "\n")
cat("Unique Pespeni SPU identifiers:", nrow(pespeni), "\n")
cat("Shared identifiers:", length(intersect(dylus$spu_id, pespeni$spu_id)), "\n")
```

# Build or recover the merged annotation table

```{r merge-annotations}
if (file.exists(eggnog_file)) {
  eggnog <- read_csv(eggnog_file, show_col_types = FALSE)

  required_eggnog <- c("gene_id", "seed_ortholog", "evalue", "score")
  if (!all(required_eggnog %in% names(eggnog))) {
    stop("eggnog_annotations.csv is missing required columns.")
  }

  eggnog_biomineralization <- eggnog |>
    mutate(seed_ortholog = normalise_spu_id(seed_ortholog)) |>
    left_join(dylus, by = c("seed_ortholog" = "spu_id")) |>
    left_join(pespeni, by = c("seed_ortholog" = "spu_id"))
} else {
  warning(
    "Using the legacy pre-merged eggnog_biomineralization.csv. ",
    "Add eggnog_annotations.csv to reproduce the joins from the original input."
  )
  eggnog_biomineralization <- read_csv(
    legacy_merged_file,
    show_col_types = FALSE
  ) |>
    mutate(seed_ortholog = normalise_spu_id(seed_ortholog))
}

eggnog_biomineralization <- eggnog_biomineralization |>
  mutate(
    gene_id = as.character(gene_id),
    evalue = suppressWarnings(as.numeric(evalue)),
    score = suppressWarnings(as.numeric(score))
  )

write_csv(
  eggnog_biomineralization,
  file.path(results_dir, "eggnog_biomineralization_merged.csv"),
  na = ""
)
```

# Select one best annotation per gene

```{r best-hit}
eggnog_best <- eggnog_biomineralization |>
  filter(!is.na(gene_id), gene_id != "") |>
  group_by(gene_id) |>
  arrange(evalue, desc(score), .by_group = TRUE) |>
  slice(1) |>
  ungroup() |>
  mutate(
    biomineralization_candidate =
      !is.na(spu_name) | !is.na(Name)
  )

cat("Unique annotated A. lixula genes:", nrow(eggnog_best), "\n")
cat(
  "Unique biomineralization candidates:",
  sum(eggnog_best$biomineralization_candidate),
  "\n"
)
```

# Functional classification

```{r functional-classes}
force_include <- "AlixG037988"

text_columns <- intersect(
  c(
    "Description", "Preferred_name", "GOs", "PFAMs", "KEGG_Pathway",
    "KEGG_ko", "Class", "Name", "Ref"
  ),
  names(eggnog_best)
)

classification_text <- apply(
  eggnog_best[, text_columns, drop = FALSE],
  1,
  function(x) paste(x, collapse = " | ")
)

biomineralization_master <- eggnog_best |>
  mutate(
    .class_text = str_to_lower(classification_text),
    Function_class = case_when(
      str_detect(
        .class_text,
        "mmp|matrix metalloproteinase|metalloprotease|peptidase m10a|astacin|adamts|serine-type endopeptidase|trypsin|chymotrypsin|peptidase inhibitor|timp"
      ) ~ "ECM remodeling / proteolysis",
      str_detect(
        .class_text,
        "sema|semaphorin|thrombospondin|integrin|cadherin|cell adhesion|scavenger receptor|srcr"
      ) ~ "ECM-associated signaling / adhesion",
      str_detect(
        .class_text,
        "calcium-activated chloride|clca|chloride channel|clcn|intracellular chloride channel|annexin|calcium ion binding|ef-hand|calmodulin|pmca|atp2b|serca|atp2a|ncx|slc8|v-atpase|atp6v|slc4|slc26|bicarbonate|carbonic anhydrase"
      ) ~ "Ion & Ca support",
      str_detect(
        .class_text,
        "collagen|c1q.*collagen|extracellular matrix|\\becm\\b|spicule|plate|spine|lectin|c-type lectin|hypp|skeletal|biomineral|calcification|proteoglycan|xylosyltransferase"
      ) ~ "ECM structure (spicule/plate/spine matrix)",
      TRUE ~ "Other / unclassified"
    ),
    Function_class = if_else(
      gene_id %in% force_include,
      "ECM structure (spicule/plate/spine matrix)",
      Function_class
    )
  ) |>
  select(-.class_text)

class_order <- c(
  "ECM structure (spicule/plate/spine matrix)",
  "ECM remodeling / proteolysis",
  "ECM-associated signaling / adhesion",
  "Ion & Ca support",
  "Other / unclassified"
)

biomineralization_candidates <- biomineralization_master |>
  filter(biomineralization_candidate | gene_id %in% force_include) |>
  mutate(Function_class = factor(Function_class, levels = class_order)) |>
  arrange(Function_class, gene_id)

write_csv(
  biomineralization_master,
  file.path(results_dir, "gene_annotations_best_hit.csv"),
  na = ""
)
write_csv(
  biomineralization_candidates,
  file.path(results_dir, "biomineralization_candidates.csv"),
  na = ""
)
writeLines(
  biomineralization_candidates$gene_id,
  file.path(results_dir, "biomineralization_candidate_genes.txt")
)
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
